Wait-free Queues with Polygarithmic Step Complexity

### 2022-07-04 article/psuedocode.tex
apply jun 30 feedback

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \tt{\sl{Tree} tree} \textsf{: A binary tree of \tt{Node}s. \tt{root} is the root node.}
%such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList} blocks}
  \textsf{ implemented with an array.}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}(-1). \tt{blocks[0]} is a block with all fields equal to zero.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Block}} 
%\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
\cmt{For a \tt{block} in a \tt{blocklist} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}. \it{put the definitions before the pseudocode}}

@@

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~index of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}
  \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}
  \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}
  \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}
  \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
\end{itemize}

@@

\it{Abbreviations:}
\begin{itemize}
 \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
 \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}, blocks[0].num\sub{x}=0})
\end{itemize}
\end{footnotesize}

@@

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
\State \tt{newBlock.element= e}
%\State \tt{b.num\sub{enq}=1}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
\State \tt{leaf.head+=1}
\State \tt{leaf.}\Call{Append}{newBlock}
\EndFunction{Enqueue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\cmt{If $E(root,b_e,i_e)$ is the response to the $D(root,b_d,i_d)$ returns \tt{<b\sub{e},i\sub{e}>}. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - (i + root.blocks[b\sub{d}-1].sum\sub{deq}) $<$ 0}}
\State \Return \tt{<-1,-->}
\Else

\State \tt{r\sub{enq}= i + root.blocks[b\sub{d}-1].sum\sub{deq} - (root.blocks[b\sub{d}-1].size - root.blocks[b\sub{d}-1].sum\sub{enq} + root.blocks[b\sub{d}-1].sum\sub{deq})}
\State \cmt{size-enqs+deqs=null deqs}
\State \Return \tt{root.\Call{DSearch}{r\sub{enq}, b\sub{d}}}
\EndIf
\EndFunction{FindResponse}

@@

\Function{Object}{Dequeue()}{}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}\cmt{\tt{r\sub{deq}} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
 \cmt{\tt{$E(root,b_{enq},i_{enq})$ is response to the $D(root,b_{deq},i_{deq})$. If the response is \tt{null} then r\sub{enq} is -1.}} \label{deqRest}
\If{\tt{r\sub{enq}==-1}}
\State \tt{output= null}
\pagebreak
\Else
\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

\end{multicols}
\end{algorithmic}
\end{algorithm}
\pagebreak

@@

\Function{void}{Propagate()}{}
\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
\State \Call{Refresh()}{} \cmt{Lemma Double Refresh}\label{secondRefresh}
\EndIf
\If{\tt{this} \bf{is not} \tt{root}}
\State \tt{parent.}\Call{Propagate()}{}
\EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.} \label{setSuper}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
\EndFor
\State \tt{\Call{CAS}{\head, h, h+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \head. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}\label{setGroup}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{np\sub{dir}= dir.num\sub{propagated}} \label{setNP}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.\size= max(root.blocks[i-1].\size + newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
\EndIf

@@

\end{multicols}
\end{algorithmic}
\end{algorithm}
\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
\Function{<int, int>}{DSearch}{\sl{int} i, \sl{int} end}

\Statex \cmt{\textmd{Searches for~the \tt{i}th enqueue of the given prefix sum of \tt{b}th block in the \tt{root}. Returns the index of the leftmost block in \tt{root.blocks} whose \sl{sum\sub{enq}} is $\geq$ \tt{i}}.}
\State \tt{start= end-1}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{r\sub{enq}}}
\State \tt{start= start-(end-start)}
\EndWhile \State
\Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, end)}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element}
\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \cmt{Search range of left child's subblocks of \tt{blocks[b]}.}
\State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.} \label{rightChildGet}
\State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
\EndIf
\EndFunction{GetEnq}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the rank of $i$th dequeue in the $b$th block of the node, among the dequeues in the root.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
\If{\tt{dir {\keywordfont is} right}}
%\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
\State \tt{i+= blocks[superBlock].sum\sub{deq-left}} \cmt{consider the dequeues from the right child}
\EndIf
\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
\EndIf
\EndFunction{Index}

@@

\Statex
%\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
\Statex \tt{blocks[]: }\sf{array of blocks}
\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
\State \Return{\tt{CAS(blocks[n], null, blk)}}
\EndFunction{TryAppend}

@@

\begin{definition}[Block]
A block is an object storing some statistics, as described in Algorithm Queue. It implicitly represents a set of operations. If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}. Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if the index of the \nf{b} is smaller than the index of the \nf{b$^\prime$}'s. 
\end{definition}

@@

\begin{lemma}[Computing Queue's Head block] \label{computeHead}
  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it. \nf{FindResponse()} returns \nf{(b, i)} which $E(root, b, i)$ is the the head of the queue in $S$. If the queue is empty in $S$ then it returns \nf{<-1,-->}.
  \end{lemma}
\begin{proof}
  The \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].\size}. It is computed while creating the block in Line~\ref{computeLength}. If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0. Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size of it.
\end{proof}
\framebox[1.1\width]{HOW?} How to prove mathematicaly that \nf{ax(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)} is the \size of the queue after the block. I can only explain it here.
%\begin{proof}
%  \nf{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \nf{head} read. If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \nf{head}. If a process dies before line 51, another process will increment \nf{head} in line 54.
%\end{proof} 
\begin{lemma}[super property]\label{superCounter}
If \nf{super[i] $\neq$ null}, then \nf{super[i]} in node \nf{n} is the index of the superblock of a block with \nf{time=i}$\pm$\nf{p}.
\end{lemma}
\pagebreak

>>>>

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
%such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList}}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Block}} 
%\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}

@@

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}
  \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}
  \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}
  \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}
  \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
\end{itemize}

@@

\it{Abbreviations:}
\begin{itemize}
 \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
 \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\})}
\end{itemize}
\end{footnotesize}

@@

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
\State \tt{newBlock.element= e}
%\State \tt{b.num\sub{enq}=1}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
\State \tt{leaf.}\Call{Append}{newBlock}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b\sub{deq}, i\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}
\State \tt{<b\sub{enq}, i\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, i\sub{deq}} 
\label{deqRest}
\If{\tt{i\sub{enq}==-1}}
\State \tt{output= null}
\pagebreak
\Else
\State \tt{output= \Call{GetEnq}{b\sub{enq}, i\sub{enq}}}
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\cmt{Returns \tt{<b\sub{e},i\sub{e}>}, if $E_{root,b_e,i_e}$ is the response to the $D_{root,b_d,i_d}$. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - i\sub{d} $<$ 0}}
\State \Return \tt{<-1,-->}
\Else
\State \tt{r\sub{enq}= i\sub{d} - root.blocks[b\sub{d}-1].size + root.blocks[b\sub{d}-1].sum\sub{enq}}
\State \Return \tt{root.\Call{DSearch}{r\sub{enq}, b\sub{d}}}
\EndIf
\EndFunction{FindResponse}

\end{multicols}
\end{algorithmic}
\end{algorithm}
%##########################################

@@

\Function{void}{Propagate()}{}
\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
\State \Call{Refresh()}{} \label{secondRefresh}
\EndIf
\If{\tt{this} \bf{is not} \tt{root}}
\State \tt{parent.}\Call{Propagate()}{}
\EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h}} \cmt{Write would work too.} \label{setSuper}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
\EndFor
\State \tt{\Call{CAS}{\head, h, h+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\cmt{Creates a block to be inserted as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's \tt{num\sub{propagated}} field. These values are used for incrementing the children's \tt{num\sub{propagated}} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}\label{setGroup}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{np\sub{dir}= dir.num\sub{propagated}} \label{setNP}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.\size= max(root.blocks[i-1].\size { }+ newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
\EndIf

@@

\end{multicols}
\end{algorithmic}
\end{algorithm}
\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e} $\geq$ 1}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} if $E_{root,e}=E_{root,b,i}$.}
\State \tt{start= end-1}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)}
\EndWhile
\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_{this,b,i}$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element}
\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{$E_{this,b,i}$ is in the left child of this node.}
\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+left.blocks[blocks[b].end\sub{left}-1].sum\sub{enq}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} 
\State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i+right.blocks[blocks[b].end\sub{right}-1].sum\sub{enq}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
\State \Return\tt{right.}\Call{GetEnq}{subBlock, i} 
\EndIf
\EndFunction{GetEnq}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
\If{\tt{dir {\keywordfont is} right}}
%\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child}
\EndIf
\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
\EndIf
\EndFunction{IndexDeq}

@@

\Statex
%\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
\Statex \tt{\sl{block[]} blocks: }\sf{array of blocks}
\Statex \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
\Statex
\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
\State \Return{\tt{CAS(blocks[n], null, blk)}}
\EndFunction{TryAppend}

@@

\begin{definition}[Block]
A block is an object storing some statistics, as described in Algorithm Queue. It implicitly represents a set of operations. If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}. Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if the index of the \nf{b} is smaller than the index of the \nf{b$^\prime$}'s. For a \tt{block} in a \tt{BlockList} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}.
\end{definition}

@@

\begin{lemma}[Computing Queue's Head block] \label{computeHead}
  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it. \nf{FindResponse()} returns \nf{(b, i)} which $E(root, b, i)$ is the the head of the queue in $S$. If the queue is empty in $S$ then it returns \nf{<-1,-->}.
  \end{lemma}
\begin{proof}
  The \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].\size}. It is computed while creating the block in Line~\ref{computeLength}. If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0. Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size  of it.

  \tt{r\sub{enq}= (i\sub{d} + root.blocks[b\sub{d}-1].sum\sub{deq}) - (root.blocks[b\sub{d}-1].size - root.blocks[b\sub{d}-1].sum\sub{enq} + root.blocks[b\sub{d}-1].sum\sub{deq})}
\end{proof}
\framebox[1.1\width]{HOW?} How to prove mathematicaly that \nf{ax(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)} is the \size of the queue after the block. I can only explain it here.
%\begin{proof}
%  \nf{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \nf{head} read. If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \nf{head}. If a process dies before line 51, another process will increment \nf{head} in line 54.
%\end{proof} 
\begin{lemma}[super property]\label{superCounter}
If \nf{super[i] $\neq$ null}, then \nf{super[i]} in node \nf{n} is the index of the superblock of a block with \nf{time=i}$\pm$\nf{p}.
\end{lemma}
\pagebreak
